---
title: "Assignment 12: Bootstrap Likelihood Ratio Test (BLRT)"
author: "Ramya Pinninti"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    highlight: monochrome
    theme: spacelab
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body {background-color: #ffffff;
      color: #000000;
      font-family: Arial, sans-serif;
      font-size: 1rem;
      line-height: 1.6;
      }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}

if (!require("VGAM")) {
  install.packages("VGAM")
  library(VGAM)
}
#### VGAM
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
 \
 
## **Assignment Objectives** 

<p>
* Enhance understanding the procedure of Bootstrap hypothesis testing.

* Implement the procedures for detecting overfitting/underfitting issues in practical applications using bootstrap likelihood ratio test.
</p>


## **Policies of Using AI Tools**

<p>
**Policy on AI Tool Use**: Please adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.
</p>

<p>
**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.
</p>




## Testing Overfitting/Underfitting

In Machine Learning and Statistics, overfitting occurs when a model is too complex and learns noise, leading to poor performance on new data, while underfitting happens when a model is too simple to capture important patterns, resulting in high errors overall; both issues are explained by the Bias–Variance Tradeoff and can cause unreliable predictions in real-world applications.


The probability density function (PDF) of the Weibull distribution is:

$$
f(t; \lambda, \beta) = \frac{\beta}{\lambda} \left( \frac{t}{\lambda} \right)^{\beta-1} \exp\left[ -\left( \frac{t}{\lambda} \right)^\beta \right], \quad t \ge 0
$$
where $\lambda > 0$ is the scale parameter (characteristic life) and $\beta > 0$ is the shape parameter.

When $\beta = 1$, the Weibull PDF simplifies to the exponential PDF:

$$
f(t; \lambda) = \frac{1}{\lambda} \exp\left( -\frac{t}{\lambda} \right)
$$
with constant hazard rate $h(t) = 1/\lambda$.


<p><font color = "darkred">**This assignment focuses on performing a hypothesis test for the shape parameter ($\beta$) of the Weibull distribution within a reliability mode**</font></p>


\begin{align}
H_0&: \beta = 1 \quad \text{(Exponential model, simpler)} \\
H_1&: \beta \neq 1 \quad \text{(Weibull model, more complex)}
\end{align}



## Steps of the BLRT


* Fit models under $H_0$ and $H_1$} to the original data, compute $\Lambda_{\text{obs}}$.

* Generate bootstrap samples under $H_0$}: 
  + Estimate parameters under $H_0$ from the original data.
  + Generate $B$ datasets by sampling from the model under $H_0$ (parametric bootstrap) or by resampling residuals/cases (nonparametric bootstrap; parametric is common for BLRT).

* For each bootstrap sample $b = 1,\dots,B$:
  + Fit $H_0$ and $H_1$ models.
  + Compute $\Lambda_b = -2[\ell_{0,b} - \ell_{1,b}]$.

* Approximate p-value:

$$
  p = \frac{1}{B} \sum_{b=1}^B I(\Lambda_b \ge \Lambda_{\text{obs}})
$$
(Often a small adjustment is made for stability: $(1 + \#\{\Lambda_b \ge \Lambda_{\text{obs}}\})/(B+1)$).



\

## **Question: Reliability Application**

<p>
A wind energy company monitors the reliability of gearboxes in 75 identical wind turbines located in a coastal wind farm. The gearbox is a critical component; its failure often leads to costly downtime and repairs. Previous studies suggest that the hazard rate (failure risk) may increase over time due to mechanical wear (fatigue, pitting, bearing degradation). Engineers want to test whether the failure time distribution follows an exponential model (constant hazard, random failures) or a Weibull model with shape parameter $k>1$ (increasing hazard, indicative of aging/degradation). The failure times (in months) are:

```
   5.2,  7.8,  9.1, 11.3, 12.5, 13.0, 14.2, 15.1, 15.9, 16.7, 17.2, 17.8, 18.4, 18.9, 
  19.3, 19.7, 20.2, 20.6, 21.0, 21.5, 21.9, 22.3, 22.7, 23.1, 23.5, 23.9, 24.3, 24.7, 
  25.1, 25.5, 25.9, 26.3, 26.7, 27.1, 27.5, 27.9, 28.3, 28.7, 29.1, 29.5, 29.9, 30.3, 
  30.7, 31.1, 31.5, 31.9, 32.3, 32.7, 33.1, 33.5, 33.9, 34.3, 34.7, 35.1, 35.5, 35.9, 
  36.3, 36.7, 37.1, 37.5, 37.9, 38.3, 38.7, 39.1, 39.5, 39.9, 40.3, 40.7, 41.1, 41.5,
  41.9, 42.3, 42.7, 43.1, 43.5
```
</p>

This assignment focuses on hypothesis $H_0: \beta = 1$ (exponential) against $H_1: \beta \neq 1$ (Weibull). This framework detects overfitting (fitting a Weibull when exponential is true) and underfitting (fitting exponential when Weibull with $\beta \neq 1$ is true). 


<p>
a). Find the MLE of the Weibull parameters $\lambda$ (scale) and $\beta$ (shape), denoted by $\hat{\lambda}$ and $\hat{\beta}$, respectively, using the `optim()` procedure. [*Hint: You should provide explicit expressions for the log-likelihood and gradient functions of the Weibull distribution parameters.*]

```{r}

# We plug in our sample data, with the goal of analyzing time-to-failure using a Weibull distribution 
#Weibull shape parameter determines failure behavior
#Here we are returning to MLE parameter estimations 
#in order to test H0: B = 1 (exponential) 
#and H1: B ≠ 1 (Weibull)

sample.data <- c(
    5.2,  7.8,  9.1, 11.3, 12.5, 13.0, 14.2, 15.1, 15.9, 16.7, 17.2, 17.8, 18.4, 18.9, 
  19.3, 19.7, 20.2, 20.6, 21.0, 21.5, 21.9, 22.3, 22.7, 23.1, 23.5, 23.9, 24.3, 24.7, 
  25.1, 25.5, 25.9, 26.3, 26.7, 27.1, 27.5, 27.9, 28.3, 28.7, 29.1, 29.5, 29.9, 30.3, 
  30.7, 31.1, 31.5, 31.9, 32.3, 32.7, 33.1, 33.5, 33.9, 34.3, 34.7, 35.1, 35.5, 35.9, 
  36.3, 36.7, 37.1, 37.5, 37.9, 38.3, 38.7, 39.1, 39.5, 39.9, 40.3, 40.7, 41.1, 41.5,
  41.9, 42.3, 42.7, 43.1, 43.5
)

#Weibull MLE estimation 
#The likelihood function measures how likely the 
#observed data is given the parameters 

#Here we maximize the log-likelihood to ensure 
#numeric stability and easier differentiation
w.loglik <- function(params, data) {
  
  shape <- params[1] #shape or Beta 
  scale <- params[2] #scale or lambda
  n <- length(data) #sample size
 
  # Weibull log-likelihood derived from PDF:
  # f(t) = (β/λ)(t/λ)^(β-1) exp(-(t/λ)^β)
  
  
  loglik <- (
         n * log(shape) - n * shape * log(scale) + 
         (shape - 1) * sum(log(data)) - 
         sum((data / scale)^shape)
          )
          
  return(loglik)
}

# The score function is the first derivative of 
#the log-likelihood. It can be used to efficiently find MLE values
#specifically, summing the logs of the pdf prevents the product of 
#probabilities from dropping to zero (underflow) during the optimization

w.score <- function(params, data) {
  
  shape <- params[1]
  scale <- params[2]
  n <- length(data)
  
  #derivative wrt shape
  grad_shape <- (
    n/shape - n * log(scale) +
    sum(log(data)) -
    sum((data / scale)^shape * log(data / scale))
  )
  
  #derivative wrt scale
  grad_scale <- (
    -(n * shape)/scale +
    (shape/scale) * sum((data / scale)^shape)
  )
  
  return(c(grad_shape, grad_scale))
}

# in a perfect Exponential distribution, the scale (lambda) is the mean
initial.params <- c(shape = 1.5, scale = mean(sample.data))

# optimization of MLE to find parameter values that 
# maximize log-likelihood 

w.mle <- optim(
  par = initial.params,
  fn = w.loglik, 
  gr = w.score, #gradient/score function created
  data = sample.data, #data passed into function
  method = "L-BFGS-B", #optimization method
  control = list(fnscale = -1, # change minimization to maximization
                 maxit = 500, # iteration cap 
                 abstol = 1e-8),
  hessian = TRUE
)
#hessian = true gives the matrix of second derivatives. 
#this allows verification of the optimization reaching a true 'peak'
#and provides a basis for standard error calculation for shape 
#and scale estimates (via Fisher information)

w.ll <- w.mle$value #maximized log likelihood

```

b). Find the MLE of the exponential parameter $\lambda$ (scale), denoted by $\hat{\lambda}$, using any procedure. [*Hint: You should provide explicit expressions for the log-likelihood and gradient functions of the exponential distribution parameters.*]

```{r}
# Under H0: β = 1, Weibull reduces to Exponential distribution:
exp.loglik <- function(lambda, data) {
  
  n <- length(data)
  
  # exponential log-likelihood
  loglik <- (
    -n * log(lambda) -
    sum(data / lambda)
  )
  
  return(loglik)
}


exp.score <- function(lambda, data) {
  
  n <- length(data)
  
  # derivative of exponential log-likelihood
  grad <- (
    -n / lambda +
    sum(data) / (lambda^2)
  )
  
  return(grad)
}



# MLE for exponential model
exp.mle <- optim(
  par = mean(sample.data),   # initial guess
  fn = exp.loglik,
  gr = exp.score,
  data = sample.data,
  method = "L-BFGS-B",
  control = list(fnscale = -1)
)


exp.ll <- exp.mle$value



```

c). Use a) and b) to perform the regular likelihood ratio $\chi^2$ test for $\beta = 1$ and report the p-value.

```{r}

#Using the regular asymptotic test as a benchmark before 
#moving onto BLRT 
#where w.ll is the maximum height of Weibull likelihood 
#exp.ll is maximum for the Exponential 

#find the difference of logs to measure the gap in explanatory power between models; Wilks' theorem proves that as n 
#grows, twice the negative log-likelihood ratio asymptotically 
#follows a Chi-Square distribution. The factor of 2 maps the 
#difference onto the Chi-Square curve 

LR.stat <- 2 * (w.ll - exp.ll)

#using function to calculate p-value and save the LRT statistic p-value
p.value.lrt <- 1 - pchisq(LR.stat, df = 1)




```


d). Use the BLRT algorithm to perform a bootstrap likelihood ratio test and report the bootstrap p-value. Note that you are expected to translate the BLRT algorithm into R code to perform the BLRT. [*Hint: The chi-square distribution should not be used in this part of the analysis.*]


```{r}

# Bootstrap Likelihood Ratio 
bootstrap_lrt_weibull <- function(data, B = 999) {
  n <- length(data)
  
  #calculate the observed statistic first 
  # LR_obs = 2 * (Weibull_Max_LogLik - Exponential_Max_LogLik)
  LR_obs <- 2 * (w.ll - exp.ll)
  
  # In an exponential distribution, lambda = mean
  lambda_null <- mean(data)
  
  # holder for statistics for the null model or case 
  LR_star <- numeric(B)
  
  # construct bootstrap loop
  for (b in 1:B) {
    
    
    # using rexp instead of rnorm because our H0 here is the Exponential distribution
    bs_sample <- rexp(n, rate = 1 / lambda_null)
    
    # fit null model or exponential
    #  can use the mean directly 
    bs_lambda_hat <- mean(bs_sample)
    bs_exp_ll <- exp.loglik(bs_lambda_hat, bs_sample)
    
    # Fit full model/Weibull
    # re-run the optimization for every bootstrap sample
    bs_w_mle <- optim(
      par = c(1.5, bs_lambda_hat), 
      fn = w.loglik,
      data = bs_sample,
      method = "L-BFGS-B",
      lower = c(0.001, 0.001),
      control = list(fnscale = -1)
    )
    bs_w_ll <- bs_w_mle$value
    
    # Calculate the LR statistic, similar to before. I am 
    # subtracting the exponential from Weibull 
    LR_star[b] <- 2 * (bs_w_ll - bs_exp_ll)
  }
  
 # p-value calculation
  # How often did the Weibull look better than the Exponential 
  # in a world where we know the Exponential was true?
  p_value <- (sum(LR_star >= LR_obs) + 1) / (B + 1)
  
  return(list(
    LR_obs = LR_obs,
    p_value = p_value,
    bootstrap_stats = LR_star
  ))
}

result <- bootstrap_lrt_weibull(sample.data, B = 999)

```

```{r}


#  results from Part A and Part B
weibull_params <- w.mle$par
exp_lambda     <- exp.mle$par


final_report <- list(
  MLE_Parameter_Estimates = data.frame(
    Model = c("Weibull (H1)", "Exponential (H0)"),
    Shape_Beta = c(round(weibull_params[1], 4), 1.0000),
    Scale_Lambda = c(round(weibull_params[2], 4), round(exp_lambda, 4))
  ),
  
  Asymptotic_LRT_Results = data.frame(
    Statistic = round(LR.stat, 4),
    P_Value = format.pval(p.value.lrt, eps = 0.001), 
    # this was 0 but shows as <0.001 in results here due to 
    #small size
    Decision = ifelse(p.value.lrt < 0.05, "Reject H0", "Fail to Reject H0")
  ),
  
  Bootstrap_LRT_Results = data.frame(
    Iterations = length(result$bootstrap_stats),
    Observed_LR = round(result$LR_obs, 4),
    P_Value = result$p_value,
    Decision = ifelse(result$p_value < 0.05, "Reject H0", "Fail to Reject H0")
  )
)
print(final_report)

```




e). Write a summary of the above analyses to address the following:

* Whether the two tests generated the same results.

* Which model is recommended for the data.

</p>

<p>

  The two tests generated similar results. They are consistent in terms of their significance. The regular LRT resulted in a p-value of 0 and the BLRT resulted in a p-value of 0.001. The Weibull Model is strongly recommended, using a significance level of 0.05. Both tests resulted in p-values far below the standard alpha level and there is enough evidence to reject H0, in this case the Exponential model, for this given set of data. 
  If this company stayed with the exponential model, underfitting would occur, leading to a situation with turbines failing sooner than the model predicts (under the null). 
  
  To summarize, both the traditional LRT and bootstrap LRT provided consistent evidence against the exponential mode. The fact that both models provided the same result increases confidence in the conclusion and recommended model. I recommend the Weibull for this data on failure time distribution of wind turbines. Using a Weibull model would allow the company to account for mechanical wear-out, leading to more accurate reliability predictions.
 
</p>




