---
title: "Assignment 5: Maximum Likelihood Estimation"
author: "Ramya Pinninti"
date: " Due: March 3, 2026"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    theme: lumen
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}
####
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
\
 
## **Assignment Objectives** 

* Comprehend the likelihood function and its properties.

* Master the maximum likelihood estimation framework and required components.

* Understand the plug-in principle underlying MLE.

* Implement maximum likelihood estimation procedures in R.

\

## **Policies of Using AI Tools**

**Policy on AI Tool Use**: Students must adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.

**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.

\

**Gamma Distribution Revisited**

Let $X$ be the two parameter Gamma random variable with density function

$$
f(x \mid \alpha, \beta) = \frac{1}{\Gamma(\alpha)\beta^\alpha} x^{\alpha-1} e^{-x/\beta}  \ \ \text{for} \ \  x > 0,
$$

where with $\alpha > 0$ (shape), $\beta>0$ (scale), and

$$
\Gamma(\alpha) = \int_{0}^{\infty} t^{\alpha-1} e^{-t} \, dt, \quad \alpha > 0.
$$

$\Gamma(x)$ can be computed in R using the `gamma()` function. The derivative of $\Gamma(x)$ are given respectively by

$$
\Gamma^\prime(z) = \Gamma (z)\psi_0(z)
$$

where $\psi_0(z) = \frac{d}{dz} \ln \Gamma{z}$. In R, the digamma function $\psi_0(z)$ is evaluated by `digamma()`.



\

<font color = "blue">This assignment focuses on finding maximum likelihood estimate of parameters $\alpha$ and $\beta$ based on a real-world application data set.</font>


\

## **Question 1: Derive gradient (first order partial derivative) of likelihood function**

Assume that $\{x_1, x_2, \cdots, x_n \} \to \text{ gamma }(\alpha, \beta)$ with density function given by

$$
f(x \mid \alpha, \beta) = \frac{1}{\Gamma(\alpha)\beta^\alpha} x^{\alpha-1} e^{-x/\beta}  \ \ \text{for} \ \  x > 0,
$$

Derive the gradient of the **log-likelihood function** with respect to the gamma distribution parameters $\alpha$ and $\beta$. To this end,


a). Write out the full **log-likelihood function** based on the given data and the density function provided above.


Assume
x₁, x₂, …, xₙ ~ Gamma(α, β)

Density:

f(x | α, β) = 1 / [Γ(α) β^α] · x^(α − 1) · exp(−x/β)

Observations are independent:

L(α, β)
= ∏ from i=1 to n
[1 / (Γ(α) β^α)] · xᵢ^(α − 1) · exp(−xᵢ/β)

Factoring step:

L(α, β)
= [1 / (Γ(α) β^α)]^n
· ∏ xᵢ^(α − 1)
· exp( − (1/β) ∑ xᵢ )

Natural log step:

ℓ(α, β)
= − n log Γ(α)
− nα log β

(α − 1) ∑ log xᵢ
− (1/β) ∑ xᵢ


b). Derive the score functions (the gradient of the log-likelihood) from the full **log-likelihood function** in part (a).

Starting:

ℓ(α, β)
= − n log Γ(α)
− nα log β

(α − 1) ∑ log xᵢ
− (1/β) ∑ xᵢ

__

d/dα ℓ(α, β)

d/dα [ − n log Γ(α) ] = − n ψ(α)

d/dα [ − nα log β ] = − n log β

d/dα [ (α − 1) ∑ log xᵢ ] = ∑ log xᵢ

Last term does not depend on α.

Simplified:

d/dα ℓ(α, β)
= − n ψ(α)
− n log β

∑ log xᵢ

Derivative with Respect to β

d/dβ ℓ(α, β)

d/dβ [ − nα log β ] = − nα / β

d/dβ [ − (1/β) ∑ xᵢ ] = (1/β²) ∑ xᵢ

Other terms do not depend on β.

Therefore:

d/dβ ℓ(α, β)
= − nα / β

(1/β²) ∑ xᵢ

∇ℓ(α, β) =

[
− n ψ(α) − n log β + ∑ log xᵢ

− nα / β + (1/β²) ∑ xᵢ
]
\


LaTeX code: 
(a) Log-Likelihood

Assume $x_1, \dots, x_n \sim \text{Gamma}(\alpha,\beta)$ with density

$$
f(x \mid \alpha,\beta)
=
\frac{1}{\Gamma(\alpha)\beta^\alpha}
x^{\alpha-1}
e^{-x/\beta}.
$$

Because the observations are independent, the likelihood is

$$
L(\alpha,\beta)
=
\prod_{i=1}^n
\frac{1}{\Gamma(\alpha)\beta^\alpha}
x_i^{\alpha-1}
e^{-x_i/\beta}.
$$

Taking the natural logarithm,

$$
\ell(\alpha,\beta)
=
- n \log \Gamma(\alpha)
- n\alpha \log \beta
+ (\alpha - 1)\sum_{i=1}^n \log x_i
- \frac{1}{\beta}\sum_{i=1}^n x_i.
$$
(b) Score Functions

Differentiate with respect to $\alpha$:

$$
\frac{d}{d\alpha}\ell(\alpha,\beta)
=
- n \psi(\alpha)
- n \log \beta
+ \sum_{i=1}^n \log x_i.
$$

Differentiate with respect to $\beta$:

$$
\frac{d}{d\beta}\ell(\alpha,\beta)
=
- \frac{n\alpha}{\beta}
+ \frac{1}{\beta^2}\sum_{i=1}^n x_i.
$$

Notes:

Used $\frac{d}{d\alpha}\log\Gamma(\alpha)=\psi(\alpha)$.  
Treated the other parameter as constant when differentiating.  
Terms not involving the differentiation variable equal zero.


## **Question 2: Birth data set**

The following R code reads in a data set containing, for each of 7 days, the lengths of time in hours spent by
women in the delivery suite while giving birth (without a ceasarian section) at John Radcliffe Hospital in
Oxford, England. The data are taken from Davison (Statistical Models. Cambridge University Press, 2003).

```
2.1, 3.4, 4.25, 5.6, 6.4, 7.3, 8.5, 8.75, 8.9, 9.5, 9.75, 10, 10.4, 10.4, 16, 19,
4, 4.1, 5, 5.5, 5.7, 6.5, 7.25, 7.3, 7.5, 8.2, 8.5, 9.75, 11, 11.2, 15, 16.5, 2.6, 
3.6, 3.6, 6.4, 6.8, 7.5, 7.5, 8.25, 8.5, 10.4, 10.75, 14.25, 14.5, 1.5, 4.7, 4.7, 
7.2, 7.25, 8.1, 8.5, 9.2, 9.5, 10.7, 11.5, 2.5, 2.5, 3.4, 4.2, 5.9, 6.25, 7.3, 7.5, 
7.8, 8.3, 8.3, 10.25, 12.9, 14.3, 4, 4, 5.25, 6.1, 6.5, 6.9, 7, 8.45, 9.25, 10.1, 
10.2, 12.75, 14.6, 2, 2.7, 2.75, 3.4, 4.2, 4.3, 4.9, 6.25, 7, 9, 9.25, 10.7
```

Assume the data are generated from a gamma distribution. The objective is to use these data and the designated algorithm to find the maximum likelihood estimates (MLEs) of the parameters $\alpha$ and $\beta$.


a). Find the MLEs of $\alpha$ and $\beta$, denoted by $\hat{\alpha}$ and $\hat{\beta}$,  using gradient-based optimization via the R function `optim()` with the gradient vector derived in Question 1.

```{r}
x <- c(
2.1, 3.4, 4.25, 5.6, 6.4, 7.3, 8.5, 8.75, 8.9, 9.5, 9.75, 10, 10.4, 10.4, 16, 19,
4, 4.1, 5, 5.5, 5.7, 6.5, 7.25, 7.3, 7.5, 8.2, 8.5, 9.75, 11, 11.2, 15, 16.5, 2.6,
3.6, 3.6, 6.4, 6.8, 7.5, 7.5, 8.25, 8.5, 10.4, 10.75, 14.25, 14.5, 1.5, 4.7, 4.7,
7.2, 7.25, 8.1, 8.5, 9.2, 9.5, 10.7, 11.5, 2.5, 2.5, 3.4, 4.2, 5.9, 6.25, 7.3, 7.5,
7.8, 8.3, 8.3, 10.25, 12.9, 14.3, 4.4, 5.25, 6.1, 6.5, 6.9, 7, 8.45, 9.25, 10.1,
10.2, 12.75, 14.6, 2, 2.7, 2.75, 3.4, 4.2, 4.3, 4.9, 6.25, 7, 9, 9.25, 10.7
)
loglik <- function(par, x) {
  alpha <- par[1]
  beta  <- par[2]
  
  n <- length(x)
  
  ll <- -n*lgamma(alpha) -
        n*alpha*log(beta) +
        (alpha - 1)*sum(log(x)) -
        (1/beta)*sum(x)
  
  return(-ll)  # negative for minimization
}
gradient <- function(par, x) {
  alpha <- par[1]
  beta  <- par[2]
  
  n <- length(x)
  
  d_alpha <- -n*digamma(alpha) -
             n*log(beta) +
             sum(log(x))
  
  d_beta  <- -(n*alpha)/beta +
              sum(x)/beta^2
  
  return(-c(d_alpha, d_beta))  # negative
}
result <- optim(
  par = c(1, 1),
  fn = loglik,
  gr = gradient,
  x = x,
  method = "BFGS"
)

result$par
```

b). Apply the method of moments to obtain estimators for $\alpha$ and $\beta$. Denote these moment estimators as $\tilde{\alpha}$ and $\tilde{\beta}$.

```{r}
xbar <- mean(x)
s2   <- var(x)

alpha_tilde <- xbar^2 / s2
beta_tilde <- s2 / xbar

alpha_tilde
beta_tilde

```

c). Conduct a brief literature review comparing the method of moments estimation (MME) and maximum likelihood estimation (MLE). Synthesize the key advantages and limitations of each, concluding with a practical recommendation.

Method of moments estimation (MME) and maximum likelihood estimation (MLE) are two ways of estimating unknown parameters in statistical models. Sample data is used in both estimation methods but the approach and statistical properties of the methods vary. 

In MME, sample moments are matched with their theoretical counterparts. Using the gamma distribution as an example, the approach leads to closed-form formulas for the parameter estimates, making interpretation easy. MME is useful for finding quick estimates. However, there could be a sacrifice in efficiency due to larger variance in the estimators. Smaller sample sizes can also cause instability.

MLE, maximum likelihood estimation, allows us to choose parameter values that maximize the likelihood function. Maximum likelihood estimators are consistent and asymptotically normal. However, MLE often necessitates numerical optimization.

MME can be a strong initial approach or serve as a starting value for optimization. If computation can be used and the model assumptions are reasonable, MLE is preferred due to stronger theoretical principles and greater efficiency.





