---
title: "Assignment 6: Constructing Likelihood Ratio Confidence Interval"
author: "Ramya Pinninti"
date: " Due: March 24, 2026"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    highlight: monochrome
    theme: spacelab
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body { background-color:white; }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}

if (!require("VGAM")) {
  install.packages("VGAM")
  library(VGAM)
}
#### VGAM
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
 \
 
## **Assignment Objectives** 

* Reinforce the likelihood concepts and MLE.

* Understand the concepts of confidence intervals.

* Master the process of finding likelihood ratio confidence interval of unknown parameter.

\

## **Policies of Using AI Tools**

**Policy on AI Tool Use**: You must adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.

**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.

\

**One Parameter Lindley Distribution**

The **Lindley distribution** is a **continuous probability distribution** proposed by D.V. Lindley in 1958. It represents a **weighted mixture** of exponential and gamma distributions, providing a flexible single-parameter model for lifetime data. 

$$
f(x;\theta) = \frac{\theta^2}{1+\theta}(1+x)e^{-\theta x}, \quad x > 0, \quad \theta > 0
$$

where $x$ = random variable (e.g., time, size, amount) and $\theta$ = shape parameter controlling the distribution.

Given an independent random sample $X_1, X_2, \dots, X_n$:

$$
L(\theta) = \prod_{i=1}^n f(x_i;\theta) = \prod_{i=1}^n \left[ \frac{\theta^2}{1+\theta} (1 + x_i) e^{-\theta x_i} \right].
$$

Let $S = \sum_{i=1}^n x_i$, $\bar{x} = S/n$,  and $C = \sum_{i=1}^n \ln(1 + x_i)$ (constant with respect to $\theta$):

$$
\ell(\theta) = \ln L(\theta) = n \ln\left( \frac{\theta^2}{1+\theta} \right) + C - \theta S.
$$

After some algebra, we obtain the closed form of the MLE of $\theta$ in the following

$$
\boxed{\hat{\theta} = \frac{1 - \bar{x} + \sqrt{\bar{x}^2 + 6\bar{x} + 1}}{2\bar{x}}}
$$

As good exercise, we can derive the following Fisher information of $\theta$:

$$
\boxed{I(\theta) = \frac{2}{\theta^2} - \frac{1}{(1+\theta)^2}}
$$


\

<font color = "blue">**This assignment focuses on constructing various confidence intervals of the shape parameter $\theta$ in the Lindley distribution.**</font>


\

## **Question: Customer Service Times (minutes)**

The customer service call duration data set originates from a major telecommunications provider in North America, operating in a highly competitive market where:

```
3.2, 5.8, 7.1, 4.5, 10.3, 6.2, 8.7, 5.1, 12.5, 6.9,
9.4, 5.7, 11.8, 4.9, 9.1, 6.5, 13.2, 7.8, 10.6, 6.1,
8.9, 5.4, 12.1, 7.3, 9.8, 5.9, 11.4, 6.8, 10.9, 7.5,
4.2, 8.3, 6.4, 14.1, 5.6, 9.7, 7.9, 11.1, 6.7, 10.2,
5.3, 8.6, 7.2, 12.9, 6.3, 9.3, 8.1, 13.7, 7.6, 10.8
```

Assuming the data follow a one-parameter Lindley distribution, construct a $95\%$ confidence interval for the parameter $\theta$ using the provided data and the specified methods. For each of the following questions, first describe your reasoning process for the analysis, then write code to perform the actual analysis. Finally, summarize the results to conclude the question.


a) Construct a **95% asymptotic confidence interval** based on the asymptotic sampling distribution of the maximum likelihood estimator (MLE) of $\theta$.

b) Construct a **95% likelihood ratio confidence interval** for $\theta$.



```{r}
# Customer service data 
service_times <- c(3.2, 5.8, 7.1, 4.5, 10.3, 6.2, 8.7, 5.1, 12.5, 6.9,
                   9.4, 5.7, 11.8, 4.9, 9.1, 6.5, 13.2, 7.8, 10.6, 6.1,
                   8.9, 5.4, 12.1, 7.3, 9.8, 5.9, 11.4, 6.8, 10.9, 7.5,
                   4.2, 8.3, 6.4, 14.1, 5.6, 9.7, 7.9, 11.1, 6.7, 10.2,
                   5.3, 8.6, 7.2, 12.9, 6.3, 9.3, 8.1, 13.7, 7.6, 10.8)

# Log likelihood and score
# To find the log likelihood for a one-parameter Lindley distribution, the natural log of the PDF is found for a sample of n observations.
lindley.loglik <- function(theta, data) {
  n <- length(data)
  # Lindley PDF: (theta^2 / (theta + 1)) * (1 + x) * exp(-theta * x)
  loglik <- 2*n*log(theta) - n*log(theta + 1) + sum(log(1 + data)) - theta*sum(data)
  return(loglik)
}

# Gradient function is the first derivative of log likelihood with respect to theta
lindley.score <- function(theta, data) {
  n <- length(data)
  grad <- (2*n)/theta - n/(theta + 1) - sum(data)
  return(grad)
}

# Optimization step for MLE
mle.result <- optim(
  par = 0.5,             
  fn = lindley.loglik,
  gr = lindley.score,
  data = service_times,
  method = "L-BFGS-B",
  lower = 0.001,         # Constraint: theta > 0
  hessian = TRUE,
  control = list(fnscale = -1, maxit = 500) 
)

theta_hat <- mle.result$par
max_loglik <- mle.result$value

# Part 1: Asymptotic CI 

# Identify pivotal quantity
# theta_hat ~ N(theta, Var_hat(theta_hat))
# Standard Error is calculated using the numerical Hessian from optim()
se <- sqrt(-1 / mle.result$hessian[1,1])

# Probability statement
alpha <- 0.05
z_val <- qnorm(1 - alpha/2) 

# rewrite inequality
asymp_ci <- c(theta_hat - z_val * se, theta_hat + z_val * se)

# Likelihood Ratio CI 

# Wilks' Theorem: LR(theta) = 2[logL(theta_hat) - logL(theta)] ~ Chi-sq(1)

crit_val <- qchisq(0.95, df = 1)
lr_target <- max_loglik - (crit_val / 2)

# Define function for determining root values
lr_root_fn <- function(theta) lindley.loglik(theta, service_times) - lr_target

# Use uniroot() to find endpoints
theta_L <- uniroot(lr_root_fn, interval = c(0.01, theta_hat))$root
theta_U <- uniroot(lr_root_fn, interval = c(theta_hat, 1.0))$root
lr_ci <- c(theta_L, theta_U)
# uniroot() was used to find the values of theta where the log-likelihood function intersects the critical threshold defined by Wilks' Theorem

# Final 
cat("MLE of Theta:", round(theta_hat, 4), "\n")
cat("95% Asymptotic CI: [", round(asymp_ci[1], 4), ",", round(asymp_ci[2], 4), "]\n")
cat("95% Likelihood Ratio CI: [", round(lr_ci[1], 4), ",", round(lr_ci[2], 4), "]\n")

```
c) Assuming the two confidence intervals above are valid, compare them in terms of performance and make a recommendation. Justify your recommendation.

The asymptotic confidence interval depends on the assumption that the sample size is large enough for the CLT to produce a normal sampling distribution. As a first-order approximation, it can be inaccurate if the likelihood surface is skewed or if the sample size is not large enough.

Likelihood ratio based intervals can produce asymmetric intervals and are preferable for small to moderate sample sizes, complex parameters, and high precision results. For non-normal data such as the Lindley distribution being modeled, this method would be preferred. As theta is greater than 0, this method is more robust as it operates directly on likelihood values, while the Wald interval could produce negative lower bounds due to high variance. 

I recommend the likelihood ratio confidence interval due to providing better coverage probability for moderate sample sizes and robustness in the case of non-normality. The Asymptotic CI is perfectly symmetric around the MLE (theta hat = 0.2191). It assumes the sampling distribution of the estimator is Normal.

However, the Likelihood Ratio CI is asymmetric. The distance from the MLE to the lower bound (0.0405) is smaller than the distance to the upper bound (0.0462). This indicates that the LR method is more sensitive to the actual skewed shape of the Lindley likelihood surface, instead of simply assuming a Normal approximation.



