---
title: "Assignment 8: Bootstrap Methods and Applications"
author: "Ramya Pinninti "
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    highlight: monochrome
    theme: spacelab
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body {background-color: #ffffff;
      color: #000000;
      font-family: Arial, sans-serif;
      font-size: 1rem;
      line-height: 1.6;
      }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}

if (!require("VGAM")) {
  install.packages("VGAM")
  library(VGAM)
}
#### VGAM
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
 \
 
## **Assignment Objectives** 

<p>
* Reinforce the understanding of Bootstrap sampling .

* Understand the bootstrap estimation: confidence interval and sampling distribution.
</p>


## **Policies of Using AI Tools**

<p>
**Policy on AI Tool Use**: Please adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.
</p>

<p>
**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.
</p>


<p>**Log-normal Distribution Revisited**</p>

<p>
If $Y = \ln(X) \sim N(\mu, \sigma^2)$, then $X$ follows a lognormal distribution $X \sim \text{Lognormal}(\mu, \sigma^2)$. The probability density is given by

$$
f(x|\mu,\sigma) = \frac{1}{x\sigma\sqrt{2\pi}} \exp\left(-\frac{(\ln x - \mu)^2}{2\sigma^2}\right), \quad x > 0
$$

After some algebra, we can express the mean and variance of the above lognormal distribution in the following


\begin{align}
\mathbb{E}[X] &= \exp\left(\mu + \frac{\sigma^2}{2}\right) \\
\text{Var}(X) &= [\exp(\sigma^2) - 1] \exp(2\mu + \sigma^2)
\end{align}


Using the relationship between normal and log-normal distribution and a sample $\{x_1, x_2, \dots, x_n\}$, the MLE estimators of $\mu$ and $\sigma^2$ are given by


\begin{align}
\hat{\mu} &= \frac{1}{n}\sum_{i=1}^n \ln(x_i) \\
\hat{\sigma}^2 &= \frac{1}{n}\sum_{i=1}^n (\ln(x_i) - \hat{\mu})^2
\end{align}


Using the plug-in principle of MLE, we have the MLE of $\mathbb{E}[X]$ and $\text{Var}(X)$ in the following

$$
\boxed{\widehat{\mathbb{E}[X]} = \exp\left(\hat{\mu} + \frac{\hat{\sigma}^2}{2}\right)}
$$

</P>



<p><font color = "blue">**This assignment focuses on constructing various bootstrap confidence intervals of the lognormal population mean $\mathbb{E}[X]$**</font></p>


\

## **Question: Trace Metal Concentrations in Soil**

<p>
Soil lead (Pb) concentrations (mg/kg) from 55 urban garden sites. Trace metals in environmental media typically follow lognormal distributions due to:

* Multiplicative processes controlling accumulation

* Positive constraints (concentrations cannot be negative)

* Right-skewed nature of contamination patterns
</p>



<p>
```
0.85, 1.23, 0.92, 3.45, 2.11, 1.56, 4.89, 2.34, 1.78, 6.72, 0.95, 1.34, 8.91, 
2.67, 1.89, 5.43, 1.12, 3.78, 2.45, 7.65, 1.05, 1.45, 12.34, 2.89, 2.01, 4.56, 
1.23, 4.32, 2.67, 9.87, 0.99, 1.56, 15.23, 3.12, 2.34, 3.89, 1.34, 5.67, 2.89, 
11.45, 1.12, 1.67, 18.90, 3.45, 2.56, 3.45, 1.45, 6.78, 3.12, 14.56, 1.23, 1.78, 
22.34, 3.78, 2.78
```
</p>

<p>

$$\boxed{\text{Instructions:}}$$

Assume the data follow a log-normal distribution. For each question in parts (b)–(d), begin by clearly explaining the reasoning behind your analytical approach. Then, develop your own R functions to implement three types of confidence intervals: the asymptotic interval, the percentile bootstrap interval, and the bias-corrected and accelerated (BCa) bootstrap interval. Use these functions to construct the required confidence intervals, and verify your results using the appropriate functions from the `boot` package.

You are encouraged to design a wrapper function that integrates all confidence interval methods, allowing users to select the desired method through an input argument.
</P>

<p>

$$\boxed{\text{Individual Questions:}}$$

a). Perform 5000 bootstrap samples to estimate the bootstrap sampling distribution of $\boxed{\widehat{\mathbb{E}[X]}}$. Display the distribution using either a histogram or a kernel density plot. Comment on the shape, variability, and any notable patterns observed in the bootstrap sampling distribution.


The bootstrap sampling distribution of the mean is usually unimodal and slightly right-skewed, reflecting the skewness of the original soil concentration data. Even with a sample size of 55, the Central Limit Theorem applies and the distribution approaches normality, but the tail of high lead concentrations can maintain a slight positive skew in the bootstrap replicates.




b). Construct a 95% bootstrap percentile confidence interval for $\mu_{LN} = \mathbb{E}[X]$.

<p>This method assumes that the sampling distribution of the statistic is a good estimate for the population distribution. It uses the $\alpha/2$ and $1-\alpha/2$ quantiles of the bootstrap distribution.
</p>

c). Construct a 95% bootstrap BCa confidence interval for $\mu_{LN} = \mathbb{E}[X]$.

<p>
Reasoning: The BCa interval is the most robust method as it adjusts for bias (if the bootstrap mean differs from the sample mean) and acceleration. This is useful for skewed environmental data.
</p>

d). Use the Central Limit Theorem to construct a 95% asymptotic confidence interval for $\mu_{LN} = \mathbb{E}[X]$.

<p>
Based on the Central Limit Theorem, for large $n$, the sampling distribution of the mean is approximately normal. 
</p>

f). Assuming the confidence intervals constructed in the previous parts are valid, evaluate their performance by comparing their widths, symmetry, stability, and sensitivity to distributional skewness. Then, provide a well‑reasoned recommendation regarding which method is most suitable for this analysis.


<p>
**Evaluation of Confidence Interval Performance**

The analysis of lead concentrations in urban garden soil requires a statistical approach that accounts for the right-skewed, log-normal nature of the data. Because trace metal accumulation is often concentrated in specific high-value areas, the choice of a confidence interval directly impacts the accuracy of the estimated population mean.
</p>

<p>
**Width and Symmetry**

The asymptotic interval is based on the Central Limit Theorem and is perfectly symmetric. This symmetry is a significant limitation for soil data. By assuming a normal distribution, this method creates a boundary that is often too low on the high end of the scale. In contrast, the percentile and BCa intervals are asymmetric. They expand further to the right, which more accurately reflects the presence of high-concentration samples in the original data set.
</p>


<p>
**Stability and Sensitivity to Skewness**

The percentile bootstrap improves upon the asymptotic method by using the actual distribution of the data rather than a theoretical curve. It is more stable when the data contains outliers. However, the BCa interval is the most sophisticated option because it addresses both bias and acceleration. Acceleration refers to the way the standard error changes relative to the parameter. For log-normal data, the BCa method is the most sensitive to skewness in a way that ensures the interval remains reliable even when the data is heavily lopsided.
</p>

<p>
**Final Recommendation**

The BCa bootstrap confidence interval is the preferred method for this study. Soil lead data is not often distributed evenly; it is defined by outliers and positive constraints. While the asymptotic method is simpler to calculate, it tends to provide an overly optimistic or narrow view of the upper limits of contamination. The BCa method provides the most rigorous correction for the natural skewness of environmental samples. This ensures a more realistic and conservative estimate of lead levels, which is necessary for making informed safety decisions regarding urban agriculture.
</p>


```{r}
lead_data <- c(0.85, 1.23, 0.92, 3.45, 2.11, 1.56, 4.89, 2.34, 1.78, 6.72, 0.95, 1.34, 8.91,
               2.67, 1.89, 5.43, 1.12, 3.78, 2.45, 7.65, 1.05, 1.45, 12.34, 2.89, 2.01, 4.56,
               1.23, 4.32, 2.67, 9.87, 0.99, 1.56, 15.23, 3.12, 2.34, 3.89, 1.34, 5.67, 2.89,
               11.45, 1.12, 1.67, 18.90, 3.45, 2.56, 3.45, 1.45, 6.78, 3.12, 14.56, 1.23, 1.78,
               22.34, 3.78, 2.78)

# ---------------------------------------------------------
# 2. CI Functions
# ---------------------------------------------------------

# a) Asymptotic (CLT)
ci_asymptotic <- function(data, conf = 0.95) {
  n <- length(data)
  x_bar <- mean(data)
  se <- sd(data) / sqrt(n)
  alpha <- 1 - conf
  z <- qnorm(1 - alpha/2)
  return(c(lower = x_bar - z * se, upper = x_bar + z * se))
}

# b) Percentile Bootstrap
ci_percentile <- function(data, R = 5000, conf = 0.95) {
  boot_means <- replicate(R, mean(sample(data, replace = TRUE)))
  alpha <- 1 - conf
  return(quantile(boot_means, probs = c(alpha/2, 1 - alpha/2)))
}

# c) BCa Bootstrap 
ci_bca_custom <- function(data, R = 5000, conf = 0.95) {
  n <- length(data)
  theta_hat <- mean(data)
  
  # Bootstrap replicates
  boot_reps <- replicate(R, mean(sample(data, replace = TRUE)))
  
  # Bias-correction (z0)
  z0 <- qnorm(sum(boot_reps < theta_hat) / R)
  
  # Acceleration 
  loo_values <- sapply(1:n, function(i) mean(data[-i]))
  avg_loo <- mean(loo_values)
  
  # Formula for acceleration 
  num <- sum((avg_loo - loo_values)^3)
  den <- 6 * (sum((avg_loo - loo_values)^2))^1.5
  a <- num / den
  
  # Adjusted alpha levels
  alpha <- 1 - conf
  z_alpha <- qnorm(alpha/2)
  z_1alpha <- qnorm(1 - alpha/2)
  
  alpha1 <- pnorm(z0 + (z0 + z_alpha) / (1 - a * (z0 + z_alpha)))
  alpha2 <- pnorm(z0 + (z0 + z_1alpha) / (1 - a * (z0 + z_1alpha)))
  
  return(quantile(boot_reps, probs = c(alpha1, alpha2)))
}

# ---------------------------------------------------------
# 3. Wrapper Function
# ---------------------------------------------------------
soil_mean_ci <- function(data, method = "bca", R = 5000, conf = 0.95) {
  switch(method,
         "asymptotic" = ci_asymptotic(data, conf),
         "percentile" = ci_percentile(data, R, conf),
         "bca"        = ci_bca_custom(data, R, conf),
         stop("Method not recognized."))
}

# ---------------------------------------------------------
# 4. Verification with boot
# ---------------------------------------------------------
if (!require("boot")) install.packages("boot")
library(boot)

# Function required by boot 
calc_mean <- function(d, i) mean(d[i])

set.seed(123)
boot_obj <- boot(data = lead_data, statistic = calc_mean, R = 5000)

# Comparing results

print(boot.ci(boot_obj, type = c("norm", "perc", "bca")))

print(soil_mean_ci(lead_data, method = "asymptotic"))
print(soil_mean_ci(lead_data, method = "percentile"))
print(soil_mean_ci(lead_data, method = "bca"))
```