---
title: "Assignment 9: Hypothesis Testing and Power and Sample size Determination"
author: "Ramya Pinninti"
output:
  html_document: 
    toc: yes
    toc_depth: 4
    toc_float: yes
    number_sections: no
    toc_collapsed: yes
    code_folding: hide
    code_download: yes
    smooth_scroll: yes
    highlight: monochrome
    theme: spacelab
  word_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    keep_md: yes
  pdf_document: 
    toc: yes
    toc_depth: 4
    fig_caption: yes
    number_sections: yes
    fig_width: 3
    fig_height: 3
editor_options: 
  chunk_output_type: inline
---

```{css, echo = FALSE}
#TOC::before {
  content: "Table of Contents";
  font-weight: bold;
  font-size: 1.2em;
  display: block;
  color: navy;
  margin-bottom: 10px;
}


div#TOC li {     /* table of content  */
    list-style:upper-roman;
    background-image:none;
    background-repeat:none;
    background-position:0;
}

h1.title {    /* level 1 header of title  */
  font-size: 22px;
  font-weight: bold;
  color: DarkRed;
  text-align: center;
  font-family: "Gill Sans", sans-serif;
}

h4.author { /* Header 4 - and the author and data headers use this too  */
  font-size: 15px;
  font-weight: bold;
  font-family: system-ui;
  color: navy;
  text-align: center;
}

h4.date { /* Header 4 - and the author and data headers use this too  */
  font-size: 18px;
  font-weight: bold;
  font-family: "Gill Sans", sans-serif;
  color: DarkBlue;
  text-align: center;
}

h1 { /* Header 1 - and the author and data headers use this too  */
    font-size: 20px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: center;
}

h2 { /* Header 2 - and the author and data headers use this too  */
    font-size: 18px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h3 { /* Header 3 - and the author and data headers use this too  */
    font-size: 16px;
    font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: navy;
    text-align: left;
}

h4 { /* Header 4 - and the author and data headers use this too  */
    font-size: 14px;
  font-weight: bold;
    font-family: "Times New Roman", Times, serif;
    color: darkred;
    text-align: left;
}

/* Add dots after numbered headers */
.header-section-number::after {
  content: ".";

body {background-color: #ffffff;
      color: #000000;
      font-family: Arial, sans-serif;
      font-size: 1rem;
      line-height: 1.6;
      }

.highlightme { background-color:yellow; }

p { background-color:white; }

}
```

```{r setup, include=FALSE}
# code chunk specifies whether the R code, warnings, and output 
# will be included in the output files.
if (!require("knitr")) {
   install.packages("knitr")
   library(knitr)
}
if (!require("pander")) {
   install.packages("pander")
   library(pander)
}
if (!require("ggplot2")) {
  install.packages("ggplot2")
  library(ggplot2)
}
if (!require("tidyverse")) {
  install.packages("tidyverse")
  library(tidyverse)
}

if (!require("plotly")) {
  install.packages("plotly")
  library(plotly)
}

if (!require("VGAM")) {
  install.packages("VGAM")
  library(VGAM)
}
#### VGAM
knitr::opts_chunk$set(echo = TRUE,       # include code chunk in the output file
                      warning = FALSE,   # sometimes, you code may produce warning messages,
                                         # you can choose to include the warning messages in
                                         # the output file. 
                      results = TRUE,    # you can also decide whether to include the output
                                         # in the output file.
                      message = FALSE,
                      comment = NA
                      )  
```
 
 \
 
## **Assignment Objectives** 

<p>
* Enhance understanding the logic and procedure of hypothesis testing .

* Implement the procedures for power and sample size calculation for basic hypothesis testing procedures using nuilt-in function and manual calculation.
</p>


## **Policies of Using AI Tools**

<p>
**Policy on AI Tool Use**: Please adhere to the AI tool policy specified in the course syllabus. The direct copying of AI-generated content is strictly prohibited. All submitted work must reflect your own understanding; where external tools are consulted, content must be thoroughly rephrased and synthesized in your own words.
</p>

<p>
**Code Inclusion Requirement**: Any code included in your essay must be properly commented to explain the purpose and/or expected output of key code lines. Submitting AI-generated code without meaningful, student-added comments will not be accepted.
</p>


## **Simple versus Composite Hypothesis*

**Simple Hypothesis**

* **Simple Hypothesis Test** is a hypothesis that completely specifies the population distribution. Mathematically, simple hypothesis fixes all parameters to specific values. For example, The simple hypotheses are: $H_0: \mu = 5$ (if $\alpha$ is known), $H_0: \mu = 100, \sigma^2 = 25$, and $H_1: p = 0.7$ for a Bernoulli distribution.

* **Example Scenario**  Test if a coin is fair:

\begin{aligned}
H_0&: p = 0.5 \quad \text{(completely specified)} \\
H_1&: p = 0.6 \quad \text{(also completely specified)}
\end{aligned}

Both are **simple** hypotheses.

**Composite Hypothesis**

* **Composite Hypothesis** is a hypothesis that does not completely specify the distribution. Mathematically, it allows a range of values for at least one parameter. For example, in one-sided: $\mu >5$; in two-sided: $\mu \le 5$.

* **Example Scenarios** 


\begin{aligned}
&H_0: \mu = 5 && \text{(simple)} \\
&H_1: \mu > 5 && \text{(composite)}
\end{aligned}



\begin{aligned}
&H_0: \mu \leq 5 && \text{(composite)} \\
&H_1: \mu > 5 && \text{(composite)}
\end{aligned}


\begin{aligned}
&H_0: \text{data follows } N(\mu, 1), \mu = 0 && \text{(simple)} \\
&H_1: \text{data follows Poisson}(\lambda) && \text{(composite – different family)}
\end{aligned}



<p><font color = "darkred">**This assignment focuses on performing performing a test of mean ($\mu$) a normal population and calculating the power and sample size based on various assumptions**</font></p>


\

## **Question: New Cholesterol Medication**

<p>
A pharmaceutical company develops **"CholestFix"** to reduce Low-Density Lipoprotein (LDL, *fat carrier that's low in density*) cholesterol. The **current standard drug** lowers LDL by an average of 25 mg/dL with a standard deviation of 15 mg/dL. A clinical trial with 5 participants were recruited in the study for three months. At the end of the study, the mean reduction is 29 mg/dL. Assume that the variance of LDL reduction of new drug is the same as that of the standard drugs.

Based on the results in the clinical trial, researchers in the company believe **CholestFix** is more effective.
</p>


<p>
a). Perform a formal hypothesis test of the researchers’ belief regarding LDL reduction, using a significance level of $\alpha = 0.05$.

**Null Hypothesis ($H_0$)**: 
$\mu = 25$ mg/dL (The mean LDL reduction is the same as the standard drug).
Alternative Hypothesis ($H_a$): $\mu \neq 25$ mg/dL (The mean LDL reduction is different).

**Calculation**:
Test Statistic ($t$): $t = \frac{29 - 25}{15/\sqrt{5}} \approx 0.5963
$Degrees of Freedom ($df$): $5 - 1 = 4$
Critical Value ($t_{crit}$): $\pm 2.776$ (for $\alpha=0.05$, two-sided).
P-value: $0.5831$

**Summary & Interpretation**: Since the $|t_{stat}| < 2.776$ and the $p$-value ($0.5831$) is  greater than $0.05$, we fail to reject the null hypothesis (H0).
Justification: Despite the observed mean of 29 being higher than 25, the sample size of 5 is too small to provide statistically significant evidence of a difference.


b). Given $n = 50, \sigma = 15, \alpha = 0.05$, and an effect size we wish to detect $\delta = 4$ mg/dL (corresponding to a reduction from 29 mg/dL to 25 mg/dL).  What is the probability that we'd detect a true improvement?

Given $n=50, \sigma=15, \alpha=0.05$, and an effect size $\delta=4$:Power ($\approx 45.6\%$): Using the non-central $t$-distribution (consistent with power.t.test in R), the probability of rejecting $H_0$ if the true mean reduction is actually $29$ is $0.4557$.

There is only a 45.6% chance of correctly detecting that CholestFix works better if the actual improvement is 4 mg/dL.
Justification: A common standard for power is 80%, meaning this design has a high risk (over 50%) of a Type II error (failing to find a difference that actually exists).

c). Determine the minimum sample size required to detect an effect size of 4 mg/dL with a power of $1 - \beta = 0.8$  and a significance level of $\alpha = 0.05$. Assume the standard deviation of LDL reduction is 15 mg/dL.

To detect a $\delta=4$ mg/dL difference with a power of $0.80$ and $\alpha=0.05$:Required $n = 113$Summary & Interpretation: The company should recruit at least 113 participants to have an 80% chance of a successful trial.
**Justification**: Increasing $n$ reduces the standard error of the mean, allowing the test to distinguish between the two distributions with greater confidence

d). **Power curve**: To assess the impact of sample size on power, we can create a power function in terms of the sample size $n$ and use the remaining information from part (b). Plot the power curve by selecting a sequence of sample sizes.
</p>

In the context of this problem, power is the probability of correctly rejecting the null hypothesis ($H_0: \mu = 25$) when the new drug actually provides a significant reduction (specifically an effect size of $\delta = 4$ mg/dL).

**The 80% Power Threshold**: The red dashed horizontal line represents the standard benchmark for clinical trials. At this level, there is an 80% chance of detecting a true treatment effect if one exists (and a 20% chance of a Type II error, or "missing" the effect).

**The Required Sample Size ($n=113$)**: The vertical green dotted line indicates the intersection where the Power Curve meets the 80% threshold. This tells us that to reliably detect an improvement of 4 mg/dL in LDL reduction with 95% confidence ($\alpha = 0.05$), this company needs a minimum of 113 participants.

**Relationship Between $n$ and Power**:The curve shows a diminishing marginal return.In the lower ranges (e.g., moving from $n=5$ to $n=50$), the power increases quickly. As the sample size grows larger, the curve flattens. Adding more participants beyond 113 continues to increase power toward 1.0 (100%), but at a slower rate. 

```{r}
# 1. Setup Parameters
delta_val <- 4    
sd_val <- 15      
sig_level <- 0.05 

# 2.  Data
n_sequence <- seq(10, 200, by = 1)
powers <- sapply(n_sequence, function(n) {
  power.t.test(n = n, 
               delta = delta_val, 
               sd = sd_val, 
               sig.level = sig_level, 
               type = "one.sample", 
               alternative = "two.sided")$power
})


plot(n_sequence, powers, type = "l", col = "blue", lwd = 2,
     main = "Power Curve for CholestFix (Two-Sided)",
     xlab = "Sample Size (n)",
     ylab = "Statistical Power",
     panel.first = grid()) #  grid lines behind the curve

abline(h = 0.80, col = "red", lty = 2)
abline(v = 113, col = "darkgreen", lty = 3) 
legend("bottomright", 
       legend = c("Power Curve", "80% Power Threshold", "Required n (113)"),
       col = c("blue", "red", "darkgreen"), 
       lty = c(1, 2, 3), 
       lwd = c(2, 1, 1),
       bg = "white")
```

<font color = "red">**Note**: For each of the questions above, write a short summary of what you observed, justify why your analysis is valid, and interpret the results.</font>

